Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite

### 2022-03-09 manuscript_v10.Rmd
manuscript v10

---
title: "Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
  - Yale: School of the Environment, Yale University, New Haven, CT
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Key Points
- The BIKER (Bayesian Inference of the $k_{600}$ Exchange Rate) algorithm predicts gas exchange velocity and $CO_2$ fluxes solely from simulated SWOT data without calibration
- BIKER is robust to measurement errors implicit in SWOT river measurements
- BIKER will allow for novel study of gas exchange spatiotemporal dynamics after SWOT's launch

## Keywords
gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite can provide new insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Similar to remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%). Finally, BIKER is robust to measurement errors implicit in the SWOT data. Upon SWOT's launch, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and near-surface turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [*Liu accepted*; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.67%- *Liu accepted*], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

@@

There is a robust existing literature exploring spatiotemporal patterns in $\delta CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a]. This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.

The structure of equation 1 necessitates that calculations of $FCO_2$ are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@hallGasExchangeStreams2020; @wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number (*Sc*) of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020]. Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models focus solely on geormorphic explanantions for variation in *k* [@hallGasExchangeStreams2020].

Applying these *k* models across watersheds, regions, or continents is called ‘upscaling.’ This upscaling allows for estimating the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

@@

Global upscalng has been performed in the literature. @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. More recently, *Liu etal accepted* performed a first assessment of monthly temporal dynamics in the global river $CO_2$ flux, though they relied on modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this.

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $k_{O_2}$ must have been to produce their 'observations'. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values against process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity. Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

We have established that literature has a reasonably good understanding of $\delta CO_2$ and a relatively poorer understanding of $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time. In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

@@

To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$. We then take these findings and explore the second question by implementing this hydraulic model, which defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements. The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that rely on in situ hydraulic measurements.

@@

![Figure 1: Conceptual overview of the BIKER algorithm. SWOT will directly measure water surface width and elevation (and thus slope), and the BIKER algorithm uses those observations to infer gas exchange velocity $k_{600}$ via a process-based model (section 2).](cache/flowchart3.jpg)

## 2 Prediciting $k_{600}$ from large-river hydraulic geometry
To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for $k_{600}$, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test. Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (section 2.3). We then empirically validate the model (section 2.3)

### 2.1 Data
We develop our hydraulic model for $k_{600}$ using multiple datasets of measured $k_{600}$ and stream hydraulics collected from the literature. In total, we obtain 763 measurements of $k_{600}$, with 701 of these measurements previously gathered by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements come from @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements. All $k_{600}$ measurements come from tracer studies and thus define $k_{600}$ at the reach scale.

@@

### 2.2 Process-based hydraulic modeling of river *k*
*k* scales with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020], and extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. The small-eddy model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface. Further, to our knowledge, this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall [equation 5- @lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993]. Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

@@

### 2.3 Deriving a large-river $k_{600}$ model
Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953]. This is a common assumption in hydraulic and geomorphic modeling of large rivers. For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

@@

### 2.4 Model validation
With equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$. Equation 7 is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

>>>>

---
title: "Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt E Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
  - Yale: School of the Environment, Yale University, New Haven, CT
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
    #  pandoc_args:
    #    - '--lua-filter=scholarly-metadata.lua'
    #    - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Key Points

-   The BIKER (Bayesian Inference of the $k_{600}$ Exchange Rate) algorithm predicts gas exchange velocity and $CO_2$ fluxes solely from simulated SWOT data without calibration
-   BIKER is robust to SWOT measurement errors
-   BIKER will allow for novel study of gas exchange spatiotemporal dynamics after SWOT's launch

## Keywords

gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry

## Abstract

Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite can provide new insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established inversion techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Similar to remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%). Finally, BIKER is robust to measurement errors implicit in the SWOT data. Upon SWOT’s launch, we suggest that BIKER will be useful in mapping global-scale fluvial gas exchange and improving CO2 emissions estimates when coupled with river CO2 models.

## 1 Introduction

Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and near-surface turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [@liuImportanceHydrologyRouting2022; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.67%- @liuImportanceHydrologyRouting2022], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

@@

There is a robust existing literature exploring spatiotemporal patterns in $\delta CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018]. This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; @liuImportanceHydrologyRouting2022], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.

The structure of equation 1 necessitates that calculations of $FCO_2$ are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@hallGasExchangeStreams2020; @wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number (*Sc*) of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers each increase gas exchange rates [@hallGasExchangeStreams2020]. Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models focus solely on physical explanations for variation in *k* [@hallGasExchangeStreams2020].

Applying these *k* models across watersheds, regions, or continents is called 'upscaling.' This upscaling allows for estimating the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

@@

Global upscalng has been performed in the literature. @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. More recently, @liuImportanceHydrologyRouting2022 performed a first upscaling assessment of monthly temporal dynamics in global river $FCO_2$, though they relied on monthly modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this. In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales.

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23--349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $k_{O_2}$ must have been to produce their 'observations'. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity. Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

We have established that literature has a reasonably good understanding of $\delta CO_2$ and a relatively poorer understanding of $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time. In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly influence fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

@@

To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river $k_{600}$ (here defined as wider than 50m to align with SWOT). We then take these findings and explore the second question by implementing this hydraulic model, which defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements. The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that rely on in situ hydraulic measurements.

@@

![Figure 1: Conceptual overview of the BIKER algorithm. 1: SWOT will directly measure water surface width and elevation (and thus slope) at a series of cross-sections within mass-conserved river reaches. 2: These hydraulics measurements are used to calculate turbulent dissipation in the river channel (section 3). 3: Turbulent dissipation is used to infer $k_{600}$ via a process-based model (section 2).](cache/maps/diagram.tif)

## 2 Predicting $k_{600}$ from large-river hydraulic geometry

To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for $k_{600}$, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test. Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (section 2.3). We then empirically validate the model (section 2.3)

### 2.1 Data

We develop our hydraulic model for $k_{600}$ using multiple datasets of measured $k_{600}$ and stream hydraulics collected from the literature. In total, we obtain 763 measurements of $k_{600}$, with 701 of these measurements previously gathered by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements come from @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements. All $k_{600}$ measurements come from tracer studies and thus define $k_{600}$ at the reach scale.

@@

### 2.2 Process-based hydraulic modeling of river *k*

*k* scales with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020], and extensive field and laboratory experiments have converged on the 'small-eddy model' for *k* as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006a; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. The small-eddy model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$ Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean [e.g. @brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface. Further, to our knowledge, this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$ Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall (equation 5). Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

@@

### 2.3 Deriving a large-river $k_{600}$ model

Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953]. This is a common assumption in hydraulic and geomorphic modeling of large rivers. For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of only `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

@@

### 2.4 Model validation

With equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$. Equation 7 is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.
